---
title: "Beginning blogdown"
subtitle: "ESM 244 - professional website term project" 
author: "Allison Horst"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview

One of your term projects for ESM 244 is to create your **professional blog** (using `blogdown` in R) that features you, your experience and qualifications, and some data science (& beyond) projects that you're proud of. The goal is to create a functional website to share with future employers, colleagues, collaborators, friends, and family to share your data science and R skills. 

### Resources 

These guidelines (and everything we'll do with `blogdown` this quarter) follow **exactly** the materials that other wonderful people have put together, including: 

- [blogdown: Creating websites with R Markdown](https://bookdown.org/yihui/blogdown/) by Yihui Xie, Amber Thomas, and Alison Hill

- Alison Hill's [Summer of Blogdown](https://summer-of-blogdown.netlify.com/)

### Set-up

#### Get `blogdown` and `hugo`

##### 1. Create a version-controlled R project

- Create a new repo on GitHub called 'my-site-initials' (e.g. my-site-ah)
- Clone the project to work locally in RStudio

##### 2. Install `blogdown` package in R (once on your computer):
```{r, eval = FALSE}
# install.packages("blogdown")
```

##### 3. Attach `blogdown` (once per session):
```{r, eval = FALSE}
library(blogdown)
```

##### 4. Install `hugo` using `blogdown` helper functions (once on your computer):
```{r, eval = FALSE}
blogdown::install_hugo()

# Check the verson of hugo:
blogdown::hugo_version()

# Note: you can also check for hugo updates if you're working on this later on:
blogdown::update_hugo()
```

From Martin Frigaard on [storybench.org](https://www.storybench.org/how-to-build-a-website-with-blogdown-in-r/): "`blogdown` is an R package that allows you to create websites from R markdown files using Hugo, an open-source static site generator written in Go and known for being incredibly fast."

### Create a website

Now that we have ourselves set-up, we can create a website using an existing hugo theme. 

**Note**: hugo themes exist in GitHub repos. The one we'll use today is [gcushen/hugo-academic](https://github.com/gcushen/hugo-academic)

**Beware**: There are many other existing hugo themes (see: https://themes.gohugo.io/), but not all of them are user-friendly, maintained, or tested with blogdown. See other recommended themes here: https://bookdown.org/yihui/blogdown/other-themes.html

I **strongly recommend** sticking with hugo-academic for your site, at least for now. 

Let's go ahead and do it by running the following: 
```{r, eval = FALSE}
new_site(theme = "gcushen/hugo-academic", 
         sample = TRUE, 
         theme_example = TRUE, 
         empty_dirs = TRUE,
         to_yaml = TRUE)
```

It may take a few moments to pull the theme. Notice that it shows up in your "Viewer" tab - and you can also have it pop-out into a browser window!  

We can also have the site immediately update after saved changes using `blogdown::serve_site()`: 

```{r, eval = FALSE}
blogdown::serve_site()
```

What happens when we `serve_site()`? 
That means that all our updates are organized and put into the `public/` folder, which is what viewers will actually see when they interact with our website online. That means: **do not change things in `public/`**, change them in the content and they will be propagated correctly into `public/`. 

### Start customizing your site!

- In the `config.toml` file, update the title
- In `config/_default/params` set a new [theme](https://sourcethemes.com/academic/themes/)
- In `content/home`:
    - Set “hero” widget to `active = false` (don't delete)
    - Set “demo” widget to `active = false` (don't delete)
    - A lot of changes made to individual sections in: content/home
- In `content/authors/index.md`, update with your info & contact information
- Update your picture: 
    - Save a professional picture (square cropped) of you as `avatar.jpg`
    - Copy and paste that into `content/authors/admin` to replace existing
    
- To remove things from navigation bar: config/_default/menus.toml (just comment out things, don't remove them to make it easier if you ever want to add it back in!), where `weight` updates the order

- Update contact information also in...config/_default/params.toml
- Link to CV: see config/_default/menus.toml option (at end):

"Link to a PDF of your resume/CV from the menu.
To enable, copy your resume/CV to `static/files/cv.pdf` and uncomment the lines below."

```{r}
# [[main]]
#    name = "CV"
#    url = "files/cv.pdf"
#    weight = 70
```


- Info on themes: https://sourcethemes.com/academic/themes/
- Make your own? https://sourcethemes.com/academic/docs/customization/#custom-theme

### Adding your first project

- Create a new folder in `content/project` called my_test_project
- Create a new R Markdown document, knit and save in the my_test_project folder you just created 
- `serve_site()` and see that your test project now shows up! 

### Publish (as an unsearchable URL) via Netlify

- Sign up for a Netlify account (https://www.netlify.com/)
- Go to https://app.netlify.com/drop
- Drop your `public/` folder where prompted
- Look, a URL is created with your site on it! 

But we don't want to have to drag & drop our new public folder every time we make changes (although you could). We want our site to auto-deploy whenever we make a stored change - in our case, when we push changes back to GitHub, and those changes will get deployed through Netlify. 

### Have Netlify get updates from GitHub

- In RStudio: stage, commit and push all updates to your GitHub repo
- Back in Netlify, click on 'Build and Deploy' and notice that you can choose to "Link site to Git". Click that.
- Choose GitHub
- You may at this point need to authenticate Netlify to communicate with GitHub > Configure Netlify on GitHub 
- Choose repository access for your website repo name

